A pipeline to analyse time-course gene expression data

### 2020-06-17 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

\begin{abstract}
The phenotypic diversity of cells is governed by a complex equilibrium between their genetic identity and their environmental interactions: Understanding the dynamics of gene expression is a fundamental question of biology.However, analysing time-course transcriptomic data raises unique challenging statistical and computational questions, requiring the development of novel methods and software. Using as case study time-course transcriptomics data from mice exposed to different strains of influenza, this workflow provides a step-by-step tutorial of the methodology used to analyse time-course data: (1) normalization of the micro-array data set; (2) differential expression analysis using functional data analysis; (3) clustering of time-course data; (4) interpreting clusters with GO term and KEGG pathway enrichment analysis.
\end{abstract}

@@

The manuscript was last compile on 2020-06-05 16:11:32.

@@

\begin{figure}

{\centering \includegraphics[width=11.53in]{images/fig1_schemas}

@@

The first steps of analysis of gene expression data is always to do
normalization and quality control checks of the data. In what follows, we show
an example of this for the influenza data using generic methods; these steps
are not in specific to time course data, but could be done for any gene expression
analysis.

@@

Before we dive into the exploratory analysis and quality control, let us
define color schemes for our data that we will use across the whole analysis.
We define color schemes for groups and time points as named vectors. We also
define a series of markers (or plotting symbols) to distinguish replicate
samples in scatter plots.

@@

\NormalTok{replicate_markers =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{15}\NormalTok{, }\DecValTok{17}\NormalTok{, }\DecValTok{19}\NormalTok{)}
\KeywordTok{names}\NormalTok{(replicate_markers) =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{3}\NormalTok{)}
\NormalTok{ann_markers =}\StringTok{ }\KeywordTok{list}\NormalTok{(}
    \DataTypeTok{Replicate=}\NormalTok{replicate_markers)}
\end{Highlighting}
\end{Shaded}

@@

\begin{figure}

{\centering \includegraphics{reports/manuscript_files/figure-latex/geneVarHist-1}

@@

\textbf{EAP}: Not sure why you hide the code for this plot, but not for others\ldots{}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{pca_data =}\StringTok{ }\KeywordTok{prcomp}\NormalTok{(}\KeywordTok{t}\NormalTok{(variance_filtered_data), }\DataTypeTok{rank=}\DecValTok{3}\NormalTok{, }\DataTypeTok{center=}\OtherTok{TRUE}\NormalTok{, }\DataTypeTok{scale=}\OtherTok{TRUE}\NormalTok{) }
\NormalTok{percent_var =}\StringTok{ }\KeywordTok{round}\NormalTok{(}\DecValTok{100} \OperatorTok{*}\StringTok{ }\KeywordTok{attr}\NormalTok{(pca_data, }\StringTok{"percentVar"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

We then plot the two first PC components, and color each sample by (1) its
condition; (2) its sampling time. We use different plotting symbols for each
replicate. We also plot the second and third components in the second column.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{))}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{4.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC1"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group[meta}\OperatorTok{$}\NormalTok{Group],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)], }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{)}
\KeywordTok{mtext}\NormalTok{(}\DataTypeTok{at=}\DecValTok{275}\NormalTok{, }\DataTypeTok{text=}\StringTok{"Colored by Group"}\NormalTok{, }\DataTypeTok{line=}\FloatTok{0.5}\NormalTok{, }\DataTypeTok{font=}\DecValTok{2}\NormalTok{, }\DataTypeTok{adj=}\FloatTok{0.5}\NormalTok{)}
\KeywordTok{legend}\NormalTok{(}\DataTypeTok{x=}\DecValTok{250}\NormalTok{, }\DataTypeTok{y=}\DecValTok{25}\NormalTok{, }\DataTypeTok{legend=}\KeywordTok{names}\NormalTok{(ann_colors}\OperatorTok{$}\NormalTok{Group),}
       \DataTypeTok{fill=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{bty=}\StringTok{"n"}\NormalTok{, }\DataTypeTok{xpd=}\OtherTok{NA}\NormalTok{, }\DataTypeTok{yjust=}\FloatTok{0.5}\NormalTok{)}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{5.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{.1}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC3"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group[meta}\OperatorTok{$}\NormalTok{Group],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
    \DataTypeTok{ylab=}\StringTok{""}\NormalTok{, }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{, }\DataTypeTok{yaxt=}\StringTok{"n"}\NormalTok{)}
\KeywordTok{axis}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DataTypeTok{labels=}\OtherTok{FALSE}\NormalTok{)}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{4.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC1"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint)],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)], }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{)}
\KeywordTok{mtext}\NormalTok{(}\DataTypeTok{at=}\DecValTok{275}\NormalTok{, }\DataTypeTok{text=}\StringTok{"Colored by Timepoint"}\NormalTok{, }\DataTypeTok{line=}\FloatTok{0.5}\NormalTok{, }\DataTypeTok{font=}\DecValTok{2}\NormalTok{, }\DataTypeTok{adj=}\FloatTok{0.5}\NormalTok{)}
\KeywordTok{legend}\NormalTok{(}\DataTypeTok{x=}\DecValTok{250}\NormalTok{, }\DataTypeTok{y=}\OperatorTok{-}\DecValTok{15}\NormalTok{, }\DataTypeTok{legend=}\KeywordTok{names}\NormalTok{(ann_colors}\OperatorTok{$}\NormalTok{Timepoint),}
       \DataTypeTok{fill=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint, }\DataTypeTok{bty=}\StringTok{"n"}\NormalTok{, }\DataTypeTok{xpd=}\OtherTok{NA}\NormalTok{, }\DataTypeTok{yjust=}\FloatTok{0.5}\NormalTok{)}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{5.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{.1}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC3"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint)],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
    \DataTypeTok{ylab=}\StringTok{""}\NormalTok{, }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{, }\DataTypeTok{yaxt=}\StringTok{"n"}\NormalTok{)}
\KeywordTok{axis}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DataTypeTok{labels=}\OtherTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/pcaPlots-1} \end{center}

Then, we plot the pearson correlation across each samples. We order the
samples by their Group (treatment) and Timepoint (the time of sampling).
Additionally, in this example, we order each treatment by strength of the
pathogenicity of the treatment: Control, Kawasaki, California, low dose of
Vietnam, then high dose of Vietnam.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Reorder the conditions such that:}
\CommentTok{#   - Control is before any influenza treatment}
\CommentTok{#   - Each treatment is ordered from low to high pathogeny}
\NormalTok{meta}\OperatorTok{$}\NormalTok{Group =}\StringTok{ }\KeywordTok{factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Group, }\KeywordTok{levels}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Group)[}\KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{4}\NormalTok{)])}

\CommentTok{# Reorder genes on condition, time, and replicate}
\NormalTok{ord =}\StringTok{ }\KeywordTok{order}\NormalTok{(}
\NormalTok{  meta}\OperatorTok{$}\NormalTok{Group,}
\NormalTok{  meta}\OperatorTok{$}\NormalTok{Timepoint,}
\NormalTok{  meta}\OperatorTok{$}\NormalTok{Replicate)}

\NormalTok{variance_filtered_data =}\StringTok{ }\NormalTok{variance_filtered_data[, ord]}
\NormalTok{data_corr =}\StringTok{ }\KeywordTok{cor}\NormalTok{(variance_filtered_data, }\DataTypeTok{method=}\StringTok{"pearson"}\NormalTok{)}
\NormalTok{data_corr_meta =}\StringTok{ }\NormalTok{meta[ord, ]}
\NormalTok{data_corr_meta}\OperatorTok{$}\NormalTok{Timepoint =}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(data_corr_meta}\OperatorTok{$}\NormalTok{Timepoint)}
\CommentTok{# use aheatmap from the NMF package for a heatmap}
\KeywordTok{aheatmap}\NormalTok{(}
\NormalTok{    data_corr,}
    \DataTypeTok{Colv=}\OtherTok{NA}\NormalTok{, }\DataTypeTok{Rowv=}\OtherTok{NA}\NormalTok{,}
    \DataTypeTok{annCol=}\NormalTok{data_corr_meta[, }\KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
    \DataTypeTok{annRow=}\NormalTok{data_corr_meta[, }\KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
    \DataTypeTok{annLegend=}\OtherTok{TRUE}\NormalTok{, }
    \DataTypeTok{annColors=}\NormalTok{ann_colors,}
    \DataTypeTok{main=}\StringTok{"Correlation plot"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}

{\centering \includegraphics{reports/manuscript_files/figure-latex/correlationPlot-1}

@@

\caption{Cross-correlation of the samples}\label{fig:correlationPlot}
\end{figure}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item
  Per-time point analysis, where we consider each time point a different
  condition and determine what genes are changing between time points, or
  between conditions at a single time-point.
\item
  Global analysis, where we consider the expression pattern globally over
  time, and consider what genes have either different patterns between
  conditions or a changing pattern (i.e.~non-constant) over time. A common
  approach first step is to fit a spline model to each gene
  \citep{storey:significance}, and then use that spline model to test for different
  kinds of differential expression across time.
\end{enumerate}

The per-time point analysis is using classical differential expression
approaches, and is often the approach advocated when dealing with small
time-course datasets, where there are only a few time points \citep{ritchie:limma, robinson:edgeR, love:moderated} . For long time-course datasets, however, a
separate test for each time-point results in creating many different tests,
for example one for every time point, the results of which are difficult to
integrate. We find in practice that the global analysis simplifies analysis
and interpretation of longer time courses data, with per-time point analysis
reserved for particularly interesting comparisons of individual time-points.

@@

\textbf{Use with \texttt{moanin}} \texttt{moanin} provides functionality for performing both of
these types of approaches, though our focus is on the global approach. In both
situations, we first need to set up a object (a \texttt{moanin\_model} object) to
hold the meta data, as well as information for fitting the spline model
(formula, number of degrees of freedom of the splines, \textless{}80\textgreater{})

We start by creating the \texttt{moanin\_model} object using the \texttt{create\_moanin\_model}
function. We need to provide two things to the function: a \texttt{data.frame} with
the metadata, and the number of degrees of freedom of the splines used in the
functional modeling. The metadata \texttt{data.frame} object should contain at least
two columns: one named \texttt{Group}, containing the treatement effect, and a
second one named \texttt{Timepoint} containing the timepoint information.

@@

The \texttt{moanin\_model} object contains a number of metadata and options used
throughout this analysis: the condition and timepoints of each samples, the
formula object used, the basis matrix, and the degrees of freedom of the model.

If no formula is provided, it will default to the following, which provides
for a different spline fit for every group (as defined by the \texttt{Group} variable
in the \texttt{meta} data.frame):
\texttt{formula\ =\ \textasciitilde{}Group:ns(Timepoint,\ df=degrees\_of\_fredoom)\ +\ Group\ +\ 0}.

The basis matrix contains the value of all of the spline basis function for
each sample of the moment. As such, replicate samples will have the same
values.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{kable}\NormalTok{(}\KeywordTok{head}\NormalTok{(moanin_model}\OperatorTok{$}\NormalTok{basis), }\DataTypeTok{digits=}\DecValTok{2}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{font_size =} \DecValTok{7}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{table}[H]
\centering\begingroup\fontsize{7}{9}\selectfont

\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r}
\hline
  & GroupM & GroupK & GroupC & GroupVL & GroupVH & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)1 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)1 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)1 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)1 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)1 & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)2 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)2 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)2 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)2 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)2 & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)3 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)3 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)3 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)3 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)3 & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)4 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)4 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)4 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)4 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)4 & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)5 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)5 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)5 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)5 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)5 & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)6 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)6 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)6 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)6 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)6\\
\hline
GSM1557140 & 0 & 1 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0\\
\hline
GSM1557141 & 0 & 1 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0\\
\hline
GSM1557142 & 0 & 1 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0\\
\hline
GSM1557143 & 0 & 1 & 0 & 0 & 0 & 0 & 0.51 & 0 & 0 & 0 & 0 & 0.04 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -0.15 & 0 & 0 & 0 & 0 & 0.35 & 0 & 0 & 0 & 0 & -0.19 & 0 & 0 & 0\\
\hline
GSM1557144 & 0 & 1 & 0 & 0 & 0 & 0 & 0.51 & 0 & 0 & 0 & 0 & 0.04 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -0.15 & 0 & 0 & 0 & 0 & 0.35 & 0 & 0 & 0 & 0 & -0.19 & 0 & 0 & 0\\
\hline
GSM1557145 & 0 & 1 & 0 & 0 & 0 & 0 & 0.51 & 0 & 0 & 0 & 0 & 0.04 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -0.15 & 0 & 0 & 0 & 0 & 0.35 & 0 & 0 & 0 & 0 & -0.19 & 0 & 0 & 0\\
\hline
\end{tabular}
\endgroup{}
\end{table}

\textbf{EAP: General comment: you need a few more print() and head() of some objects
to demonstrate what the output looks like}

@@

\texttt{moanin} provides a simple interface to perform a timepoint by timepoint
differential expression analysis. Comparison between groups is traditionally
done by defining the group comparisons (called \texttt{contrasts} in linear models)
as a linear combination of the coefficients of the model. Comparing groups
within each timepoint can create many contrasts, and thus \texttt{moanin} provides
functionality to create these contrasts in an automatic way, and then calls
\texttt{limma} \citep{ritchie:limma} on the set of contrasts provided. By default,
\texttt{moanin} expects RNA-Seq contact counts, and will estimate voom weights.

Here, we show an example where we define our contrasts to be the difference
between the control mouse (``M'') and the mouse infected with the high dose of
the influenza strain A/Vietnam/1203/04 (H5N1) (``VL'') for each time point, but
the function works with any form contrasts \citep{ritchie:limma}.

@@

This creates a vector of contrasts to be tested, one for each timepoint, in the format required by \texttt{limma}:

\begin{tabular}{l}
\toprule
Contrasts\\
\midrule
M.0-VL.0\\
M.3-VL.3\\
M.6-VL.6\\
M.9-VL.9\\
M.12-VL.12\\
\addlinespace
M.18-VL.18\\
M.24-VL.24\\
M.30-VL.30\\
M.36-VL.36\\
M.48-VL.48\\
\addlinespace
M.60-VL.60\\
M.72-VL.72\\
M.120-VL.120\\
M.168-VL.168\\
\bottomrule
\end{tabular}

@@

\begin{tabular}{lrrrrrr}
\toprule
  & M.0-VL.0\_pval & M.0-VL.0\_qval & M.0-VL.0\_lfc & M.3-VL.3\_pval & M.3-VL.3\_qval & M.3-VL.3\_lfc\\
\midrule
NM\_009912 & 0.273 & 0.475 & 2.202 & 0.767 & 0.876 & 1.158\\
NM\_008725 & 0.851 & 0.924 & 2.028 & 0.139 & 0.302 & 1.457\\
NM\_007473 & 0.088 & 0.219 & 0.678 & 0.917 & 0.959 & 0.764\\
ENSMUST00000094955 & 0.008 & 0.037 & 1.186 & 0.935 & 0.969 & -0.633\\
NM\_001042489 & 0.419 & 0.621 & 1.212 & 0.761 & 0.872 & 0.599\\
\addlinespace
NM\_008159 & 0.123 & 0.278 & 1.102 & 0.117 & 0.268 & 0.737\\
NM\_001013813 & 0.057 & 0.159 & 0.630 & 0.171 & 0.348 & 0.975\\
AK039774 & 0.015 & 0.059 & 0.637 & 0.062 & 0.171 & 0.971\\
NM\_013782 & 0.875 & 0.938 & 0.594 & 0.000 & 0.001 & 1.180\\
NM\_028622 & 0.767 & 0.876 & 1.648 & 0.946 & 0.974 & -1.865\\
\bottomrule
\end{tabular}

@@

Let's look at the distribution of genes found differentially expressed per
week between control and each of the influenza strains.

>>>>

\begin{abstract}
The phenotypic diversity of cells is governed by a complex equilibrium between their genetic identity and their environmental interactions: Understanding the dynamics of gene expression is a fundamental question of biology. However, analysing time-course transcriptomic data raises unique challenging statistical and computational questions, requiring the development of novel methods and software. Using as case study time-course transcriptomics data from mice exposed to different strains of influenza, this workflow provides a step-by-step tutorial of the methodology used to analyse time-course data: (1) quality control and normalization of the data set; (2) differential expression analysis using functional data analysis; (3) clustering of time-course data; (4) interpreting clusters with GO term and KEGG pathway enrichment analysis.
\end{abstract}

@@

\textbf{EAP}: I don't know that describing normalization as part of our tutorial makes sense.
I've tried to change the text.

The manuscript was last compile on 2020-06-17 05:37:22.

@@

\begin{figure}[H]

{\centering \includegraphics[width=.5\textwidth,]{images/fig1_schemas}

@@

\hypertarget{overview-of-the-data}{%
\subsection{Overview of the data}\label{overview-of-the-data}}

@@

Before we dive into the exploratory analysis and quality control, let us
define color schemes for our data that we will use across the whole analysis.
We define color schemes for groups and time points as named vectors. We also
define a series of markers (or plotting symbols) to distinguish replicate
samples in scatter plots. We also reorder the factor \texttt{Group} which describes the treatments so that the treatments are order from low to high pathogeny (with Control being first).

@@

\NormalTok{replicate_markers =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{15}\NormalTok{, }\DecValTok{17}\NormalTok{, }\DecValTok{19}\NormalTok{)}
\KeywordTok{names}\NormalTok{(replicate_markers) =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{3}\NormalTok{)}
\NormalTok{ann_markers =}\StringTok{ }\KeywordTok{list}\NormalTok{(}
    \DataTypeTok{Replicate=}\NormalTok{replicate_markers)}

\CommentTok{# Reorder the conditions such that:}
\CommentTok{#   - Control is before any influenza treatment}
\CommentTok{#   - Each treatment is ordered from low to high pathogeny}
\NormalTok{meta}\OperatorTok{$}\NormalTok{Group =}\StringTok{ }\KeywordTok{factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Group, }\KeywordTok{levels}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Group)[}\KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{4}\NormalTok{)])}
\end{Highlighting}
\end{Shaded}

The first steps of analysis of gene expression data is always to do
normalization and quality control checks of the data. However, in what follows, we do not show the steps for normalization, as these are specific to the platform (microarray); the code for the normalization can be found in as a supplemental file. \textbf{EAP}: perhaps a citation to review papers for normalization on microarray and mRNA-Seq.

Instead, we focus on common steps for exploratory analysis of the data, including for the purpose of quality control. These steps
are not specific to time course data, but could be applied for any gene expression
analysis. For this reason, we will not print the detailed code that is needed for this part of the analysis; interested readers can examine the \texttt{rmarkdown} code that accompanies this workflow.

@@

\begin{figure}[H]

{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/geneVarHist-1}

@@

@@

In Figure \ref{fig:pcaPlots}, we visualize the data by plotting the PC components, with samples colored by either its
condition (top row) or its sampling time (bottom row) and each replicate a different symbol. We can see a large difference in later time points.

\begin{figure}[H]

{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/pcaPlots-1}

@@

\caption{Plot of PC components: PC2 vs PC1 (right) and PC2 vs PC3 (left). Samples are colored by condition (top row) and sampling time (bottom row).}\label{fig:pcaPlots}
\end{figure}

We also plot in Figure \ref{fig:correlationPlot} the pearson correlation between each sample as a heatmap diagram (using the function \texttt{aheatmap} in \texttt{NMF}). We order the
samples by their Group (treatment) and Timepoint (the time of sampling).

\begin{figure}[H]

{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/correlationPlot-1}

\caption{Heatmap of the correlation between samples}\label{fig:correlationPlot}
\end{figure}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item
  Per-time point analysis, where we consider each time point a different
  condition and determine what genes are changing between specific time points, or
  between conditions at a single time-point.
\item
  Global analysis, where we consider the expression pattern globally over
  time, and consider what genes have either different patterns between
  conditions or a changing pattern (i.e.~non-constant) over time. A common
  approach first step is to fit a spline model to each gene
  \citep{storey:significance}, and then use that spline model to test for different
  kinds of differential expression across time.
\end{enumerate}

The per-time point analysis is using classical differential expression
approaches, and is often the approach advocated when dealing with small
time-course datasets, where there are only a few time points \citep{ritchie:limma, robinson:edgeR, love:moderated}. For long time-course datasets, however, a
separate test for each time-point results in creating many different tests,
for example one for every time point, the results of which are difficult to
integrate. We find in practice that the global analysis simplifies analysis
and interpretation of longer time courses data, with per-time point analysis
reserved for particularly interesting comparisons of individual time-points.

@@

\textbf{DE Analysis in \texttt{moanin}} Our package \texttt{moanin} provides functionality for performing both of
these types of approaches, though our focus is on the global approach, specifically by fitting spline models to the genes.

In both
situations, we first need to set up a object (a \texttt{moanin\_model} object) to
hold the meta data, as well as information for fitting the spline model
(formula, number of degrees of freedom of the splines, \ldots{}). The \texttt{moanin\_model} object
will contain information used
throughout this analysis, in particular the condition and timepoints of each sample and the
the basis matrix for fitting splines models.

We start by creating the \texttt{moanin\_model} object using the \texttt{create\_moanin\_model}
function. We need to provide two things to the function: a \texttt{data.frame} with
the metadata, and the number of degrees of freedom of the splines to be used in the
functional modeling. The metadata \texttt{data.frame} object should contain at least
two columns: one named \texttt{Group}, containing the treatement effect, and a
second one named \texttt{Timepoint} containing the timepoint information.

@@

The main operation of the \texttt{create\_moanin\_model} function, in addition to holding the necessary meta data of the samples,
is to create a basis matrix for the splines fit. This matrix gives the evaluation of all of the spline basis functions for
each sample (as such, replicate samples will have the same
values). By default, \texttt{create\_moanin\_model} will create the spline basis functions which will lead to a different spline fit for every group (as defined by the \texttt{Group} variable
in the \texttt{meta} data.frame). This is done by the following \texttt{R} formula syntax:

\texttt{formula\ =\ \textasciitilde{}Group:ns(Timepoint,\ df=degrees\_of\_fredoom)\ +\ Group\ +\ 0}.

Alternatively, the user can provide a formula of their own, or simply provide the basis matrix themselves.

We can see this information when we print the object:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{print}\NormalTok{(moanin_model)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## moanin_object on 209 samples containing the following information:
## 1) Meta data with 5 variables:
## [1] ""            "Group"       "Replicate"   "Timepoint"   "WeeklyGroup"
## 2) Basis matrix with 35 basis functions
## Basis matrix was constructed with the following formula
## ~Group + Group:splines::ns(Timepoint, df = degrees_of_freedom) + 0 
## Where degrees_of_freedom=6 
## To access these objects use:
##  <object_name>$degrees_of_freedom
##  <object_name>$basis
##  <object_name>$meta
##  <object_name>$formula
\end{verbatim}

@@

\texttt{moanin} provides a simple interface to perform a timepoint by timepoint
differential expression analysis. Comparison between groups is traditionally
done by defining the group comparisons (called \texttt{contrasts} in linear models)
as a linear combination of the coefficients of the model. Comparing groups
within each timepoint can create many contrasts, and thus \texttt{moanin} provides
functionality to create these contrasts in an automatic way, and then calls
\texttt{limma} \citep{ritchie:limma} on the set of contrasts provided. By default,
\texttt{moanin} expects RNA-Seq gene counts, and will estimate voom weights.

Here, we show an example where we create contrasts that will be be the difference
between the control mouse (``M'') and the mouse infected with the high dose of
the influenza strain A/Vietnam/1203/04 (H5N1) (``VL'') for each time point (but
the function works with any form contrasts \citep{ritchie:limma}).

@@

This creates a character vector of contrasts to be tested, one for each timepoint, in the format required by \texttt{limma}:

\begin{verbatim}
## [1] "M.0-VL.0"   "M.3-VL.3"   "M.6-VL.6"   "M.9-VL.9"   "M.12-VL.12"
\end{verbatim}

\textbf{EAP}: I think it's better here to actually show the output, rather than a table, so you can see what the contrasts actually look like. I like to reserve table format for actual results, rather than code-like input

@@

\begin{tabular}{lrrr}
\toprule
  & M.0-VL.0\_pval & M.0-VL.0\_qval & M.0-VL.0\_lfc\\
\midrule
NM\_009912 & 0.273 & 0.475 & 2.202\\
NM\_008725 & 0.851 & 0.924 & 2.028\\
NM\_007473 & 0.088 & 0.219 & 0.678\\
ENSMUST00000094955 & 0.008 & 0.037 & 1.186\\
NM\_001042489 & 0.419 & 0.621 & 1.212\\
\addlinespace
NM\_008159 & 0.123 & 0.278 & 1.102\\
NM\_001013813 & 0.057 & 0.159 & 0.630\\
AK039774 & 0.015 & 0.059 & 0.637\\
NM\_013782 & 0.875 & 0.938 & 0.594\\
NM\_028622 & 0.767 & 0.876 & 1.648\\
\bottomrule
\end{tabular}

@@

In Figure \ref{fig:timeSignifPerTime}, we show the distribution of genes found differentially expressed per
week between control and each of the influenza strains. Such an analysis can demonstrate some general trends, with clearly more genes
being differentially expressed at later time points, and the Vietnam high-dose
showing perhaps an earlier onset than the low-dose.
